---
title: "DataExploration"
author: "Vix Talbot and Haziel Garcia Sanchez"
date: "2024-02-23"
output: html_document
---
Data Description:

The College data set contains a number of variables for 777 different universities and colleges in the US. The variables are

• Private : Public/private indicator
• Apps : Number of applications received
• Accept : Number of applicants accepted
• Enroll : Number of new students enrolled
• Top10perc : New students from top 10 % of high school class • Top25perc : New students from top 25 % of high school class 
• F.Undergrad : Number of full-time undergraduates
• P.Undergrad : Number of part-time undergraduates
• Outstate : Out-of-state tuition
• Room.Board : Room and board costs
• Books : Estimated book costs
• Personal : Estimated personal spending
• PhD : Percent of faculty with Ph.D.’s
• Terminal : Percent of faculty with terminal degree • S.F.Ratio : Student/faculty ratio
• perc.alumni : Percent of alumni who donate
• Expend : Instructional expenditure per student
• Grad.Rate : Graduation rate

```{r setup, include=FALSE}
# load and examine the data set
library(ISLR)
library(ISLR2)

data(College)
head(College)

# View(College)

```



```{r}
# part a of the project questions

college_lm <- lm(Apps ~ ., data = College)

summary(college_lm)

# checking git stuff

```